---
title: "Lognormal race with unknown winner"
author: "Bruno Nicenboim"
date: "6/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)
```

```{r}
set.seed(42)
library(ggplot2)
library(tidyr)
library(dplyr)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = 3)
```

I'm here trying to simulate the outcome of two processes, one is the time-out (`to`) that goes down according to the practice law, and another one `proc` that depends on a predictor `X`. The two processes follow lognormal distributions with the same variance, and the fastest one is the observed one. Importantly, I can't know directly which is the one that finished first in each trial.

```{r}
N <- 5000
X <- rnorm(N)
Int_to <- log(1000) ##log(1000) = 6.9
Int_proc <- 6
r <- .0005
#effect of something:
mu_proc <- Int_proc + X *.1
#practice:
mu_to <- Int_to + -r*(1:N)
sigma <- .5
proc <- rlnorm(N,mu_proc,sigma)
to <- rlnorm(N,mu_to,sigma)
data <- tibble(trial=1:N,X, 
               proc=proc, 
               to=to, 
               RT = pmin(proc, to), 
               ## IRL, I don't know which one actually wins:
               winner_proc = proc< to)
```

It will look like this (the crosses are the observations):
```{r}
data %>% gather(key="type",value="T", proc, to) %>%
    ggplot( aes(x=trial, y=T, color = type)) + geom_point(alpha=.5) +
    geom_point(aes(y=RT), size=2, color= "black", alpha=.1, shape =3)
```

This is the proportion of times `proc` won and was observed:
```{r}
mean(proc < to)
mean(data$winner_proc)
```

Because these are two lognormal distributions, I can also calculate the proportions of the mixture like this, for every trial:
```{r}
plnorm(1, mu_proc - mu_to, sqrt(2) * sigma) %>% head()
```

And on average:
```{r}
plnorm(1, mu_proc - mu_to, sqrt(2) * sigma) %>% mean()
```

## Stan model of the race (a lognormal race model with an unknown winner)

I try to fit this in stan with quite informative priors:

```{r stan_code, tidy = TRUE, comment="", echo=FALSE}
cat(readLines("race_proc.stan"), sep = "\n")  
```

```{r mixture, cache = TRUE, results = "hide"}
data_simple <- as.list(data) %>% c(N=nrow(data))
out_simple <- stan("race_proc.stan", data=data_simple, chains =6,iter=1000)
```

It doesn't work:
```{r}
print(out_simple, pars= c("Int_proc","Int_to","slope", "r", "sigma", "P_proc_avg"), probs = c(0.025,  0.50,  0.975))
```

It seems that there are a ton of modes, but they don't overlap with the correct true values of the parameters!
```{r}
traceplot(out_simple, pars= c("Int_proc","Int_to","slope", "r", "sigma", "P_proc_avg"))
```

These are the summaries of the individual chains, they are actually too precise:
```{r}
summary(out_simple,  pars= c("Int_proc","Int_to","slope", "r", "sigma", "P_proc_avg"), probs = c(0.025,  0.50,  0.975))
```

## Stan model of a lognormal race model, assuming I know the winner (which I don't)

If I know which process is the fastest one, this is simply a lognormal race model, and it's quite easy to fit:


```{r stan_code_known, tidy = TRUE, comment="", echo=FALSE}
cat(readLines("race_proc_known.stan"), sep = "\n")  
```

```{r, cache = TRUE, results = "hide"}
data_simple <- as.list(data) %>% c(N=nrow(data))
out_simple_known <- stan("race_proc_known.stan", data=data_simple, chains =6,iter=1000)
```

It works, but I'm using data that I shouldn't have:
```{r}
print(out_simple_known, pars= c("Int_proc","Int_to","slope", "r", "sigma", "P_proc_avg"), probs = c(0.025,  0.50,  0.975))
```

